---
title: "PCA"
author: "Deviprasad Saka"
date: "2024-03-03"
output: html_document
---

```{r}

library(MASS)
library(factoextra)
library(ggplot2)


# Define the file path
file_path <- file.path("D:", "Multivariate Analysis", "Datasets", "Body Measurements _ original_CSV.csv")

# Read the CSV file
body <- read.csv(file_path)
```

```{r}
dim(body)
```

```{r}
str(body)
```

```{r}
summary(body)
```
Summary:

Gender: This variable is categorical and has two unique values, "Male" and "Female". The number of observations (length) is 716.

Age: This variable represents the age of the individuals in the dataset. The minimum age is 1, the maximum age is 68, and the mean age is approximately 15.34. The median age is 11, which indicates that half of the individuals are younger than 11 and half are older.

Head Circumference, Shoulder Width, Chest Width, Belly, Waist, Hips, Arm Length, Shoulder to Waist, Waist to Knee, Leg Length, Total Height: These variables represent various body measurements. For each variable, the summary includes the minimum, maximum, mean, median, and quartile values (1st quartile, median, and 3rd quartile). These values provide information about the distribution of each body measurement in the dataset.

Overall, this summary provides a quick overview of the distribution of each numerical variable in the dataset, which can be useful for understanding the characteristics of the data and identifying potential outliers.
```{r}
# Exclude categorical data from column 1 and Run PCA
body_sample <- body[, -1]
# Run PCA
body_pca <- prcomp(body_sample, scale= TRUE)
```

```{r}
#Summary of analysis
summary(body_pca)
```
summary:
Proportion of Variance: This row represents the proportion of total variance explained by each principal component. It indicates the contribution of each principal component to the overall variability of the data. For example, PC1 explains 44.45% of the total variance, PC2 explains 11.56%, and so on.

Cumulative Proportion: This row represents the cumulative proportion of total variance explained by each principal component. It indicates the cumulative contribution of principal components up to a certain point. For example, the cumulative proportion of variance explained by PC1 and PC2 is 56.02% (44.45% + 11.56%), and so on.

```{r}
#Elements of PCA object
names(body_pca)
```


```{r}
# std deviation of components
body_pca$sdev
```

```{r}
#Eigen vectors
body_pca$rotation
```


```{r}
#std deviation and mean of variables
body_pca$center
body_pca$scale
```

```{r}
# Principal component scores
body_pca$x
```

```{r}
# Scree plot of Variance
fviz_eig(body_pca, addlabels = TRUE, ylim = c(0, 50))
```


```{r}
# Biplot with default settings
fviz_pca_biplot(body_pca)
```


```{r}
# Biplot with Labeled variables
fviz_pca_biplot(body_pca, label="var")
```


```{r}
# Biplot with colored groups
fviz_pca_biplot(body_pca, label="var", habillage = body$Gender)
```

```{r}
# Biplot with customized Colored Groups and Variables
fviz_pca_biplot(body_pca, label = "var", habillage = body$Gender) + 
  scale_color_manual(values = c("Male" = "orange", "Female" = "purple"))
```




```
